---
title: "Plastic Pollution and its Economic Drivers"
author: "Jonathan Flores Martinez, Avery Stewart, Justin Mai"
theme: cerulean
embed-resources: true
format: 
  dashboard:
    orientation: rows
nav-buttons:
      - icon: github
        href: https://github.com/jonflores27/stat541-project-2
        aria-label: GitHub
logo: plastics.jpeg
execute: 
  echo: false
  warning: false
  message: false
---

# Main Question

```{r}
#| label: Libraries
#| message: false

library(tidyverse)
library(gt)
library(httr)
library(stringr)
library(tidyjson)
library(jsonlite)
library(ggridges)
library(RColorBrewer)
library(viridis)
library(ggtext)
library(broom)
library(scales)
library(car)
library(plotly)
library(MASS)

plastics <- readr::read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/main/data/2021/2021-01-26/plastics.csv')

plastics_clean <- plastics |> 
  mutate(across(empty:pvc, ~replace_na(., 0)),
         total = rowSums(across(c(empty:pvc))),
         .before = num_events)
final_merged <- readRDS("full_table")
```



## Row - Research Question {height="10%"}

::: {.card}
**Research Question**

How are demographic, economic, and geographic characteristics of countries associated with differences in plastic collected during cleanup audits?
:::

## Row {height="90%"}

### Column {width="55%"}

#### Row {height="60%}
::: panel-tabset
## GDP per Capita Groups

```{r}
#| label: Vizualization-3

final_merged |> 
  filter(year == 2019) |> 
  group_by(country) |> 
  summarize(unknown = sum(empty),
            hdpe = sum(hdpe),
            ldpe = sum(ldpe),
            other = sum(o),
            pet = sum(pet),
            pp = sum(pp),
            ps = sum(ps),
            pvc = sum(pvc),
            total = sum(total),
            gdp_per_capita = mean(gdp_per_capita),
            hdi = mean(HDI),
            .groups = "drop") |> 
  mutate(income_group = case_when(
    gdp_per_capita < 1500 ~ "Low Income",
    gdp_per_capita > 12000 ~ "High Income",
    TRUE ~ "Middle Income")) |> 
  pivot_longer(
    cols = unknown:pvc,
    names_to = "plastic_type",
    values_to = "waste") |> 
  filter(waste <= 3500) |> 
  
  ggplot(mapping = aes(x = waste,
                       y = plastic_type,
                       fill = income_group)) +
  geom_density_ridges(alpha = 0.7) +
  scale_fill_manual(values = c(
  "High Income"   = "#58b4f5",
  "Middle Income" = "#a8d8ea",
  "Low Income"    = "#56e8c8")) +
  labs(
    title = "Plastic Waste Type Distribution of Countries by GDP per Capita",
    subtitle = "Divided into <span style='color: #58b4f5;'>High Income</span>, 
      <span style='color: #a8d8ea;'>Middle Income</span>,
    and <span style='color: #56e8c8;'>Low Income</span>",
    x = "Waste Amount",
    y = "",
    alt = "Density ridge plot showing the density of plastic waste audited
    for multiple types of plastic. Colored by GDP per capita category,
    High, Middle, and Low") +
  theme_dark() +
  theme(plot.subtitle =  element_markdown(),
        legend.position = "none") 

```

## GDP vs Plastic Pollution

```{r}
#| label: Vizualization-2
#| warning: false

final_merged |>
  mutate(region = countrycode::countrycode(country,
            origin = "country.name",
            destination = "continent"),
         total_normalized = total / num_events, 
         gdp_trillions = gdp / 1e6) |>
  filter_out(country == "EMPTY",
           total == 0,
           is.na(total),
           is.na(gdp_trillions),
           gdp_trillions == 0, 
           is.na(region)) |>
  group_by(country, region, year) |>
  summarize(total_plastic = sum(total_normalized, na.rm = T), 
            gdp_trillions = mean(gdp_trillions, na.rm = T), 
            Population = mean(Population, na.rm = T)) |>
  ggplot(aes(x = gdp_trillions, y = total_plastic, size = Population, 
             color = region)) +
    geom_point(alpha = 0.7) +
    scale_size(range = c(.1, 15), name ="Population") +
  labs(color = "Region",
       x = "GDP (Trillions)", 
       title = "Plastic Pollution Found in Audits vs. GDP for 2019 and 2020",
       subtitle = "Total Plastic per Cleanup Event",
       y = "",
       alt = "Bubble chart showing plastic items found per cleanup event vs GDP in trillions,
       colored by region and sized by population") +
  coord_cartesian(ylim = c(0, 26000),
                  xlim = c(0,25)) + 
  scale_color_discrete(na.translate = FALSE) +
  scale_x_sqrt() +
  facet_wrap(~ year) +
  theme_light() + 
  scale_color_manual(values = c(
  "Africa"   = "#E69F00", 
  "Americas" = "#56B4E9",
  "Asia"     = "#009E73", 
  "Europe"   = "#F0E442", 
  "Oceania"  = "#0072B2"
))

```
:::

#### Row 


```{r}
the_7 <- final_merged |> 
  filter(!parent_company %in% c("Grand Total", 
                                "Unbranded", 
                                "null", 
                                "NULL", 
                                "Assorted")) |> 
  mutate(country = str_to_title(country)) |>
  group_by(country, parent_company) |> 
  summarize(total = sum(total), .groups = "drop") |> 
  group_by(country) |> 
  slice_max(order_by = total, n = 7, with_ties = FALSE) |> 
  ungroup()
```




```{r}
#| execute: false
#| include: false

the_7 |> 
  filter(country == "Argentina") |> 

ggplot(mapping = aes(x = reorder(parent_company, total), y = total)) +
  geom_col() +
  labs(title = "Top 7 Parent Companies in Argentina",
       x = "Parent Company",
       y = "Total") +
  theme_minimal()
```



```{r}

countries <- unique(the_7$country)

fig <- plot_ly()

for (cty in countries) {

  df <- the_7 |>
    filter(country == cty) |>
    arrange(total) |>
    mutate(parent_company = factor(parent_company, levels = parent_company))

  fig <- fig |>
    add_bars(data = df,
             x = ~parent_company,
             y = ~total,
             name = cty,
             visible = FALSE)
}

buttons <- lapply(seq_along(countries), function(i) {

  df <- the_7 |>
    filter(country == countries[i]) |>
    arrange(total)

  vis <- rep(FALSE, length(countries))
  vis[i] <- TRUE

  list(method = "update",
       args = list(list(visible = vis),
                   list(title = paste("Top 7 Plastic Waste Producing Companies in",
                                      countries[i]),
                        xaxis = list(tickmode = "array",
                                     tickvals = 0:nrow(df),
                                     ticktext = df$parent_company))),
       label = countries[i])})

fig <- fig |>
  layout(updatemenus = list(
    list(type = "dropdown", buttons = buttons)),
    title = paste("Top 7 Plastic Waste Producing Companies in", countries[1]),
    xaxis = list(title = "Parent Company"),
    yaxis = list(title = "Total Plastic Audited"),
    showlegend = FALSE)

fig
```


### Column - Model {width="45%"}

```{r}
#| label: Data Model Output1
#| warning: false

model_df <- final_merged |>
  group_by(country, year) |>
  summarise(
    total_plastic = sum(total, na.rm = TRUE),
    Population_millions = first(na.omit(Population)) / 1000000,
    pop_density = first(na.omit(pop_density)),
    gdp_billions = first(na.omit(gdp)) / 1000000000,
    region = first(na.omit(region)),
    .groups = "drop"
  ) |>
  mutate(
    region = as.factor(region),
    log_total_plastic = log(total_plastic)
  ) 

fit_6 <- lm(
  log_total_plastic ~ Population_millions + pop_density + gdp_billions + region,
  model_df
)

tbl <- tidy(fit_6, conf.int = TRUE) |>
  mutate(
    Predictor = dplyr::recode(term,
      `(Intercept)` = "Intercept",
      `Population_millions` = "Population (millions)",
      `pop_density` = "Population Density",
      `gdp_billions` = "GDP (billions)"
    ),
    Predictor = sub("^region", "Region: ", Predictor)
  ) |>
  dplyr::select(
    Predictor,
    Estimate = estimate,
    `Std. Error` = std.error,
    `t value` = statistic,
    `p value` = p.value,
    `95% CI Lower` = conf.low,
    `95% CI Upper` = conf.high
  )

tbl |>
  gt() |>
  tab_header(
    title = md("**Linear Regression Predicting Logged Plastic Collected by Country, Year**"),
    subtitle = "Response: log(total_plastic)"
  ) |>
  cols_label(
    Predictor = md("**Predictor**"),
    Estimate = md("**Estimate**"),
    `Std. Error` = md("**Std. Error**"),
    `t value` = md("**t value**"),
    `p value` = md("**p value**"),
    `95% CI Lower` = md("**95% CI Lower**"),
    `95% CI Upper` = md("**95% CI Upper**")
  ) |>
  fmt_number(
    columns = c(Estimate, `Std. Error`, `t value`, `95% CI Lower`, `95% CI Upper`),
    decimals = 4
  ) |>
  fmt_number(columns = `p value`, decimals = 3) |>
  tab_source_note(
    source_note = md(
       "The response variable was transformed as log(total_plastic) to meet linear regression assumptions"
    )
  )
```





# Nigeria

## Row - Why Nigeria? {height="25%"}

::: {.card title="Why Nigeria?"}


Nigeria ranks among the highest contributors to plastic waste in our dataset when summarized over all years. This made it a compelling country for more exploration and understanding other factors that contribute to the high levels of plastic pollution. By focusing on Nigeria, we aim to uncover how environmental conditions and governmental infrastructure may affect the country's plastic waste output.

:::

## Row - Plot of Nigeria

```{r}
#| warning: false

library(sf)
library(tidyverse)
library(leaflet)

my_shapefile <- st_read("nga_admin1.geojson", quiet = TRUE)
nigera <- read_csv("data.csv")

nigera_clean <- nigera |>
  filter(State != "TOTAL") |>
  mutate(
    State = case_when(
      State == "Abuja" ~ "Federal Capital Territory",
      State == "Akwa-Ibom" ~ "Akwa Ibom",
      State == "Cross-river" ~ "Cross River",
      State == "Bayesa" ~ "Bayelsa",
      TRUE ~ State
    )
  )

plot_this <- my_shapefile |>
  left_join(nigera_clean, by = c("adm1_name" = "State")) |>
  mutate(
    plastic_kg_per_person_year = (Ty / Population) / 11 * 1000
  )

centroids <- st_centroid(plot_this)

pal <- colorBin(
  palette = "Blues",
  domain = plot_this$plastic_kg_per_person_year,
  bins = c(0, 25, 50, 75, 100, 150, 250),
  na.color = "grey80"
)

leaflet(plot_this) |>
  addProviderTiles(providers$CartoDB.PositronNoLabels) |>
  addPolygons(
    fillColor = ~pal(plastic_kg_per_person_year),
    fillOpacity = 0.8,
    color = "black",
    weight = 1,
    popup = ~ifelse(
      is.na(plastic_kg_per_person_year),
      paste0("<b>", adm1_name, "</b><br>No data available"),
      paste0(
        "<b>", adm1_name, "</b>",
        "<br>Total plastic (Ty): ", round(Ty, 0),
        "<br>Population: ", Population,
        "<br>Plastic per person: ",
        round(plastic_kg_per_person_year, 2), " kg/year"
      )
    )
  ) |>
  addLabelOnlyMarkers(
    data = centroids,
    label = ~adm1_name,
    labelOptions = labelOptions(
      noHide = TRUE,
      direction = "center",
      style = list(
        "font-weight" = "bold",
        "color" = "black",
        "font-size" = "9px",
        "padding" = "2px"
      )
    )
  ) |>
  addLegend(
    pal = pal,
    values =  ~plastic_kg_per_person_year[!is.na(plastic_kg_per_person_year)],
    title = "Plastic waste (kg per person per year)",
    position = "bottomright"
  ) |>
  addControl(
    "No data = grey states",
    position = "bottomright"
  ) |>
  addControl(
    "Data available for 28 states (2007–2017)",
    position = "topright"
  )

```