---
title: "Plastic Pollution and its Economic Drivers"
author: "Jonathan Flores Martinez, Avery Stewart, Justin Mai"
theme: cerulean
embed-resources: true
format: 
  dashboard:
    orientation: rows
nav-buttons:
      - icon: github
        href: https://github.com/jonflores27/stat541-project-2
        aria-label: GitHub
logo: plastics.jpeg
execute: 
  echo: false
  warning: false
  message: false
server: shiny
---

# The World

```{r}
#| label: Libraries
#| message: false
#| context: setup

library(tidyverse)
library(gt)
library(httr)
library(stringr)
library(tidyjson)
library(jsonlite)
library(ggridges)
library(RColorBrewer)
library(viridis)
library(ggtext)
library(broom)
library(scales)
library(car)
library(plotly)
library(MASS)
library(shiny)
library(sf)
library(leaflet)
library(lme4)
library(lmerTest)
library(broom.mixed)


plastics <- readr::read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/main/data/2021/2021-01-26/plastics.csv')

plastics_clean <- plastics |> 
  mutate(across(empty:pvc, ~replace_na(., 0)),
         total = rowSums(across(c(empty:pvc))),
         .before = num_events)
final_merged <- readRDS("full_table")

state_gdp <- read_csv("nigeria_state_gdp.csv")
census <- read_csv("nigera-2020.csv")


state_gdp_clean <- state_gdp |>
  mutate(State = str_remove(State, " State"))

census_clean <- census |> 
  mutate(State = str_remove(State, " State"),
         `Population 2020 projected` = `Population 2020 projected`/1000000)


my_shapefile <- st_read("nga_admin1.geojson", quiet = TRUE)
nigeria      <- read_csv("data.csv")
gdp_data     <- read_csv("nigeria_state_gdp.csv")

nigeria_clean <- nigeria |>
  filter(State != "TOTAL") |>
  mutate(
    State = case_when(
      State == "Abuja"        ~ "Federal Capital Territory",
      State == "Akwa-Ibom"   ~ "Akwa Ibom",
      State == "Cross-river" ~ "Cross River",
      State == "Bayesa"      ~ "Bayelsa",
      TRUE ~ State
    )
  )

plot_this <- my_shapefile |>
  left_join(nigeria_clean, by = c("adm1_name" = "State")) |>
  mutate(plastic_kg_per_person_year = (Ty / Population) / 11 * 1000)


plot_this2 <- my_shapefile |>
  left_join(state_gdp_clean, by = join_by(adm1_name == State)) |>
  mutate(gdp_bil_usd = `GDP (bil. US$)`)

plot_this3 <- my_shapefile |> 
  left_join(census_clean, by = join_by(adm1_name == State))
  


centroids <- st_centroid(plot_this)
centroids3 <- st_centroid(plot_this3)




first_nonmissing <- function(x) {
  x <- x[!is.na(x)]
  if (length(x) == 0) NA else x[1]
}
model_df <- final_merged |>
  group_by(country, year) |>
  summarise(
    total_plastic = sum(total, na.rm = TRUE),
    Population_millions = first_nonmissing(Population) / 1000000,
    pop_density = first_nonmissing(pop_density),
    gdp_billions = first_nonmissing(gdp) / 1000000000,
    region = first_nonmissing(region),
    .groups = "drop"
  ) |>
  mutate(
    country = factor(country),
    year = factor(year),
    region = factor(region),
    log_total_plastic = log(total_plastic),
    log_population = log(Population_millions),
    log_pop_density = log(pop_density),
    log_gdp_billions = log(gdp_billions)
  ) 




```



## Row - Research Question {height="10%"}

::: {.card}
**Research Question**

How are demographic, economic, and geographic characteristics of countries associated with differences in plastic collected during cleanup audits?
:::

## Row {height="90%"}

### Column {width="60%"}

#### Row {height="45%"}
::: panel-tabset
## GDP vs Plastic Pollution
::: {.card title="The Big Picture"}
GDP is the one of the most popular measures of a country's economic health. In this plot, we can see how that surface level measurement relates to the plastic audited by region through the world.

```{r}
#| label: Vizualization-2
#| warning: false

final_merged |>
  mutate(region = countrycode::countrycode(country,
            origin = "country.name",
            destination = "continent"),
         total_normalized = total / num_events, 
         gdp_trillions = gdp / 1e6) |>
  filter_out(country == "EMPTY",
           total == 0,
           is.na(total),
           is.na(gdp_trillions),
           gdp_trillions == 0, 
           is.na(region)) |>
  group_by(country, region, year) |>
  summarize(total_plastic = sum(total_normalized, na.rm = T), 
            gdp_trillions = mean(gdp_trillions, na.rm = T), 
            Population = mean(Population, na.rm = T)) |>
  ggplot(aes(x = gdp_trillions, y = total_plastic, size = Population, 
             color = region)) +
    geom_point(alpha = 0.7) +
    scale_size(range = c(.1, 15), name ="Population") +
  labs(color = "Region",
       x = "GDP (Trillions)", 
       title = "Plastic Pollution Found in Audits vs. GDP for 2019 and 2020",
       subtitle = "Average Amount of Plastic Audited per Cleanup Event in a Country",
       y = "",
       alt = "Bubble chart showing plastic items found per cleanup event vs GDP in trillions,
       colored by region and sized by population") +
  coord_cartesian(ylim = c(0, 26000),
                  xlim = c(0,25)) + 
  scale_color_discrete(na.translate = FALSE) +
  scale_x_sqrt() +
  facet_wrap(~ year) +
  theme_light() + 
  scale_color_manual(values = c(
  "Africa"   = "#E69F00", 
  "Americas" = "#56B4E9",
  "Asia"     = "#009E73", 
  "Europe"   = "#F0E442", 
  "Oceania"  = "#0072B2"
))

```
:::

## GDP vs Plastic Types
::: {.card title="Platic Types by GDP Categories"}
The plastic type audited in a country may be affected by the country's economic health. This plot differentiates this health by the GDP per Capita metric. Once again often used to denote economic health.
```{r}
#| label: Vizualization-3

final_merged |> 
  filter(year == 2019) |> 
  group_by(country) |> 
  summarize(unknown = sum(empty),
            hdpe = sum(hdpe),
            ldpe = sum(ldpe),
            other = sum(o),
            pet = sum(pet),
            pp = sum(pp),
            ps = sum(ps),
            pvc = sum(pvc),
            total = sum(total),
            gdp_per_capita = mean(gdp_per_capita),
            hdi = mean(HDI),
            .groups = "drop") |> 
  mutate(income_group = case_when(
    gdp_per_capita < 1500 ~ "Low Income",
    gdp_per_capita > 12000 ~ "High Income",
    TRUE ~ "Middle Income")) |> 
  pivot_longer(
    cols = unknown:pvc,
    names_to = "plastic_type",
    values_to = "waste") |> 
  filter(waste <= 3500) |> 
  
  ggplot(mapping = aes(x = waste,
                       y = plastic_type,
                       fill = income_group)) +
  geom_density_ridges(alpha = 0.7) +
  scale_fill_manual(values = c(
  "High Income"   = "#E69F00",
  "Middle Income" = "#009E73",
  "Low Income"    = "#0072B2")) +
  labs(
    title = "Plastic Waste Type Distribution of Countries by GDP per Capita",
    subtitle = "Divided into <span style='color: #E69F00;'>High Income</span>, 
      <span style='color: #009E73;'>Middle Income</span>,
    and <span style='color: #0072B2;'>Low Income</span>",
    x = "Plastic Waste Units Audited",
    y = "",
    alt = "Density ridge plot showing the density of plastic waste audited
    for multiple types of plastic. Colored by GDP per capita category,
    High, Middle, and Low") +
  theme_minimal() +
  theme(plot.subtitle =  element_markdown(),
        legend.position = "none") 

```
:::


## Waste Producers
::: {.card title="Economic Drivers = Plastic Waste Generators???"}
The presence of large corporations withing a country may also affect the plastic produced. This table explores which companies produce the most plastic waste that was audited within a country.

```{r}
the_7 <- final_merged |> 
  filter(!parent_company %in% c("Grand Total", 
                                "Unbranded", 
                                "null", 
                                "NULL", 
                                "Assorted")) |> 
  mutate(country = str_to_title(country),
         country = case_when(country == "United Kingdom Of Great Britain & Northern Ireland" ~ "UK & Northern Ireland",
                             country == "Taiwan_ Republic Of China (Roc)" ~ "Taiwan",
                             .default = country)) |>
  group_by(country, parent_company) |> 
  summarize(total = sum(total), .groups = "drop") |> 
  group_by(country) |> 
  slice_max(order_by = total, n = 7, with_ties = FALSE) |> 
  ungroup()
```

```{r}


#COPY THIS TABLE FOR CHANGE

countries <- unique(the_7$country)

fig <- plot_ly()

for (cty in countries) {

  df <- the_7 |>
    filter(country == cty) |>
    arrange(total) |>
    mutate(parent_company = factor(parent_company, levels = parent_company))

  fig <- fig |>
    add_bars(data = df,
             x = ~total,
             y = ~parent_company,
             name = cty,
             visible = (cty == countries[1]),
             orientation = "h")
}

buttons <- lapply(seq_along(countries), function(i) {
  df <- the_7 |>
    filter(country == countries[i]) |>
    arrange(total)
  
  vis <- rep(FALSE, length(countries))
  vis[i] <- TRUE

  list(method = "update",
       args = list(list(visible = vis),
                   list(title = paste(
                     "Top 7 Plastic Waste Producing Companies in", 
                     countries[i]),
                     yaxis = list(title = ""),
                     xaxis = list(title = "Total Plastic Audited",
                                  autorange = TRUE))),
       label = countries[i])
})

fig <- fig |>
  layout(margin = list(t = 120, l = 220),
         updatemenus = list(list(type = "dropdown",
                                 active = 0,
                                 buttons = buttons,
                                 x = 0,
                                 y = 1.25,
                                 xanchor = "left",
                                 yanchor = "top",
                                 pad = list(r = 10, t = 10))),
         title = list(text = paste("Top 7 Plastic Waste Producing Companies in",
                                   countries[1]),
                      y = .93),
         yaxis = list(title = ""),
         xaxis = list(title = "Total Plastic Audited", autorange = TRUE),
         showlegend = FALSE)

fig

# Ik it comes out overlapping here but its fine when it renders.
```

:::

## Environmental Metrics
::: {.card title="Environmental Metrics for 5 Most Populated Countries"}

```{r}
#| label: Table Reserch Question 2
#| tbl-cap: "Summary of Plastic Compared to Environmental Indicators for the 5 Most Populated Countries"

final_merged |>
  group_by(country, year) |>
  summarise(
    total_plastic = sum(total),
    avg_plastic = mean(total),
    footprint_per_capita = first(EFConsPerCap[!is.na(EFConsPerCap)]),
    biocapacity_per_capita = first(BiocapPerCap[!is.na(BiocapPerCap)]),
    hdi = first(HDI[!is.na(HDI)]),
    population = first(Population[!is.na(Population)]),
    .groups = "drop"
  ) |>
  filter_out(country == "EMPTY") |>
  filter(country %in% c("India",
                        "China",
                        "United States of America",
                        "Indonesia", 
                        "Nigeria")) |> 
  mutate(country = str_to_title(country)) |> 
  gt() |>
  tab_header(
    title = "Country-Year Summary of Plastics and Environmental Data"
  ) |>
  cols_label(
    country = "Country",
    year = "Year",
    total_plastic = "Total Plastic Items Collected",
    avg_plastic = "Average Plastic Items Collected",
    footprint_per_capita = "Ecological Footprint per Capita",
    biocapacity_per_capita = "Biocapacity per Capita",
    hdi = "HDI",
    population = "Population"
  ) |>
  fmt_number(
    columns = c(total_plastic, avg_plastic, population),
    decimals = 0
  ) |>
  fmt_number(
    columns = c( footprint_per_capita, biocapacity_per_capita, hdi),
    decimals = 2
  )
```

:::

:::



### Column - Model {width="40%"}

::: {.card title="Our Model"}

This mixed effects model examines how population, population density, GDP, year, and geographic region are associated with the amount of plastic collected in cleanup audits. The response variable is the logged total number of plastic items collected for each country/year.

Population, population density, and GDP are also log transformed so the model is more stable rather than the raw units. Region and year are included as fixed effects, while country is included as a random effect to account for repeated observations from the same country across years.

Only log population is statistically significant in the mixed model, suggesting that countries with larger populations tend to have higher audited plastic totals, while population density, GDP, year, and region do not show significant associations after accounting for country level variation.

:::
```{r}
fit_final_maybe <- lmer(
  log_total_plastic ~
    log_population +
    log_pop_density +
    log_gdp_billions +
    year +
    region +
    (1 | country),
  data = model_df,
  REML = TRUE
)

anova(fit_final_maybe) |>
  as.data.frame() |>
  tibble::rownames_to_column("Predictor") |>
  mutate(
    Predictor = dplyr::recode(
      Predictor,
      log_population = "Log Population",
      log_pop_density = "Log Population Density",
      log_gdp_billions = "Log GDP (billions)",
      year = "Year",
      region = "Region"
    )
  ) |>
  gt() |>
  tab_header(
    title = md("**Type III ANOVA for Final Mixed Model**"),
    subtitle = "Satterthwaite's method"
  ) |>
  fmt_number(
    columns = where(is.numeric),
    decimals = 3
  ) |>
  tab_style(
    style = list(
      cell_text(weight = "bold")
    ),
    locations = cells_body(
      rows = Predictor == "Log Population"
    )
  )
```






# Nigeria

## Row - Why Nigeria? {height="25%"}

::: {.card title="Why Nigeria?"}


Nigeria ranks among the highest contributors to plastic waste in our dataset when summarized over all years. This made it a compelling country for more exploration and understanding other factors that contribute to the high levels of plastic pollution. By focusing on Nigeria, we aim to uncover how environmental conditions and governmental infrastructure may affect the country's plastic waste output.

:::

## Row - Plot of Nigeria {height="75%"}


### Column {width="20%"}

```{r}
selectInput(
  "metric",
  "Select metric to display:",
  choices = c(
    "Plastic waste (kg/person/year)" = "plastic",
    "GDP (billion US$)"              = "gdp",
    "Population 2020 (millions)"     = "population"
  ),
  selected = "plastic"
)
```

### Column {width="80%"}

```{r}
leafletOutput("map", height = "500px")
```

```{r}
#| context: server

output$map <- renderLeaflet({

  if (input$metric == "plastic") {

    pal <- colorBin(
      palette  = "Blues",
      domain   = plot_this$plastic_kg_per_person_year,
      bins     = c(0, 25, 50, 75, 100, 150, 250),
      na.color = "grey80"
    )

    leaflet(plot_this) |>
      addProviderTiles(providers$CartoDB.PositronNoLabels) |>
      addPolygons(
        fillColor   = ~pal(plastic_kg_per_person_year),
        fillOpacity = 0.8,
        color       = "black",
        weight      = 1,
        popup = ~ifelse(
          is.na(plastic_kg_per_person_year),
          paste0("<b>", adm1_name, "</b><br>No data available"),
          paste0(
            "<b>", adm1_name, "</b>",
            "<br>Total plastic (Ty): ", round(Ty, 0),
            "<br>Population: ", Population,
            "<br>Plastic per person: ", round(plastic_kg_per_person_year, 2), " kg/year"
          )
        )
      ) |>
      addLabelOnlyMarkers(
        data         = centroids,
        label        = ~adm1_name,
        labelOptions = labelOptions(
          noHide    = TRUE,
          direction = "center",
          style     = list(
            "font-weight" = "bold",
            "color"       = "black",
            "font-size"   = "9px",
            "padding"     = "2px"
          )
        )
      ) |>
      addLegend(
        pal      = pal,
        values   = ~plastic_kg_per_person_year[!is.na(plastic_kg_per_person_year)],
        title    = "Plastic waste (kg per person per year)",
        position = "bottomright"
      ) |>
      addControl("No data = grey states",               position = "bottomright") |>
      addControl("Data available for 28 states (2007–2017)", position = "topright")

  } else if (input$metric == "gdp") {

    pal <- colorBin(
      palette  = "Greens",
      domain   = plot_this$gdp_bil_usd,
      bins     = c(0, 5, 10, 20, 40, 75, 110),
      na.color = "grey80"
    )

    leaflet(plot_this2) |>
      addProviderTiles(providers$CartoDB.PositronNoLabels) |>
      addPolygons(
        fillColor   = ~pal(gdp_bil_usd),
        fillOpacity = 0.8,
        color       = "black",
        weight      = 1,
        popup = ~ifelse(
          is.na(gdp_bil_usd),
          paste0("<b>", adm1_name, "</b><br>No data available"),
          paste0(
            "<b>", adm1_name, "</b>",
            "<br>GDP: $", round(gdp_bil_usd, 2), " billion USD"
          )
        )
      ) |>
      addLabelOnlyMarkers(
        data         = centroids,
        label        = ~adm1_name,
        labelOptions = labelOptions(
          noHide    = TRUE,
          direction = "center",
          style     = list(
            "font-weight" = "bold",
            "color"       = "black",
            "font-size"   = "9px",
            "padding"     = "2px"
          )
        )
      ) |>
      addLegend(
        pal      = pal,
        values   = ~gdp_bil_usd[!is.na(gdp_bil_usd)],
        title    = "GDP (billion $US)",
        position = "bottomright"
      ) |>
      addControl("No data = grey states", position = "bottomright")
  }
  
  else if (input$metric == "population") {

    pal <- colorBin(
      palette  = "Purples",
      domain   = plot_this$`Population 2020 projected`,
      bins     = c(0, 3, 6, 7, 8, 9, 10),
      na.color = "grey80"
    )

    leaflet(plot_this3) |>
      addProviderTiles(providers$CartoDB.PositronNoLabels) |>
      addPolygons(
        fillColor   = ~pal(`Population 2020 projected`),
        fillOpacity = 0.8,
        color       = "black",
        weight      = 1,
        popup = ~ifelse(
          is.na(`Population 2020 projected`),
          paste0("<b>", adm1_name, "</b><br>No data available"),
          paste0(
            "<b>", adm1_name, "</b>",
            "<br>Populqtion in Millions", round(`Population 2020 projected`, 2)
          )
        )
      ) |>
      addLabelOnlyMarkers(
        data         = centroids3,
        label        = ~adm1_name,
        labelOptions = labelOptions(
          noHide    = TRUE,
          direction = "center",
          style     = list(
            "font-weight" = "bold",
            "color"       = "black",
            "font-size"   = "9px",
            "padding"     = "2px"
          )
        )
      ) |>
      addLegend(
        pal      = pal,
        values   = ~`Population 2020 projected`[!is.na(`Population 2020 projected`)],
        title    = "Population in Millions",
        position = "bottomright"
      ) |>
      addControl("No data = grey states", position = "bottomright")
  }
  
})
```