---
title: "Plastic Pollution and its Economic Drivers"
author: "Jonathan Flores Martinez, Avery Stewart, Justin Mai"
theme: cerulean
embed-resources: true
format: 
  dashboard:
    orientation: rows
nav-buttons:
      - icon: github
        href: https://github.com/jonflores27/stat541-project-2
        aria-label: GitHub
logo: plastics.jpeg
execute: 
  echo: false
  warning: false
  message: false
---

# Main Question

```{r}
#| label: Libraries
#| message: false

library(tidyverse)
library(gt)
library(httr)
library(stringr)
library(tidyjson)
library(jsonlite)
library(ggridges)
library(RColorBrewer)
library(viridis)
library(ggtext)
library(broom)
library(scales)
library(car)

plastics <- readr::read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/main/data/2021/2021-01-26/plastics.csv')

plastics_clean <- plastics |> 
  mutate(across(empty:pvc, ~replace_na(., 0)),
         total = rowSums(across(c(empty:pvc))),
         .before = num_events)
final_merged <- readRDS("full_table")
```



## Row - Research Question {height="11%"}

::: card
**Research Question**

How are demographic, economic, and geographic characteristics of countries associated with differences in plastic collected during cleanup audits?
:::

## Row {height="89%"}

### Column {width="50%"}

#### Row {height="50%"}


::: panel-tabset
## Environmental Metrics

```{r}
#| label: Table Reserch Question 2
#| tbl-cap: "Summary of Plastic Compared to Environmental Indicators for the 5 Most Populated Countries"

final_merged |>
  group_by(country, year) |>
  summarise(
    total_plastic = sum(grand_total),
    avg_plastic = mean(grand_total),
    footprint_per_capita = first(EFConsPerCap[!is.na(EFConsPerCap)]),
    biocapacity_per_capita = first(BiocapPerCap[!is.na(BiocapPerCap)]),
    hdi = first(HDI[!is.na(HDI)]),
    population = first(Population[!is.na(Population)]),
    .groups = "drop"
  ) |>
  filter_out(country == "EMPTY") |>
  filter(country %in% c("India",
                        "China",
                        "United States of America",
                        "Indonesia", 
                        "Nigeria")) |> 
  mutate(country = str_to_title(country)) |> 
  gt() |>
  tab_header(
    title = "Country-Year Summary of Plastics and Environmental Data"
  ) |>
  cols_label(
    country = "Country",
    year = "Year",
    total_plastic = "Total Plastic Items Collected",
    avg_plastic = "Average Plastic Items Collected",
    footprint_per_capita = "Ecological Footprint per Capita",
    biocapacity_per_capita = "Biocapacity per Capita",
    hdi = "HDI",
    population = "Population"
  ) |>
  fmt_number(
    columns = c(total_plastic, avg_plastic, population),
    decimals = 0
  ) |>
  fmt_number(
    columns = c( footprint_per_capita, biocapacity_per_capita, hdi),
    decimals = 2
  )
```

## Employment Sectors

```{r}
#| label: Research question 2 table 2
#| tbl-cap: Percent of Plastic Types Compared to Largest Indurtry Sectors for the 5 Most Populated Countries

rate_ify2 <- function(df, numerators, denominator) {
  stopifnot(is.data.frame(df))
  
  df |>
    mutate(across(.cols = {{ numerators }},
                  .fns = ~ .x / {{ denominator }},
                  .names = "{.col}"))
}

final_merged |> 
  group_by(country) |> 
  summarise(hdpe = sum(hdpe),
            ldpe = sum(ldpe),
            o = sum(o),
            pet = sum(pet),
            pp = sum(pp),
            ps = sum(ps),
            pvc = sum(pvc),
            total_waste = sum(total),
            industry = mean(employment_industry),
            service = mean(employment_services),
            ag = mean(employment_agriculture),
            .groups = "drop") |>
  mutate(
    largest_sector = case_when(
      industry >= service & industry >= ag ~ "Industry",
      service >= industry & service >= ag ~ "Service",
      ag >= industry & ag >= service ~ "Agriculture"),
    .after = total_waste ) |> 
  filter(country %in% c("India",
                        "China",
                        "United States of America",
                        "Indonesia", 
                        "Nigeria")) |>
  rate_ify2(numerator = hdpe:pvc,
           denominator = total_waste) |> 

  gt() |> 
  tab_header(
    title = "Proportions of Plastic Types Compared to Employment Sectorization",
    subtitle = "Data from 2019 and 2020 Featuring the 5 Most Populated Countries") |> 
  fmt_percent(
    columns = hdpe:pvc,
    decimals = 2) |> 
  tab_style(style = cell_text(weight = "bold"),
            locations = cells_column_labels()) |>
  cols_label(country = "Country", 
             total_waste = "Total Count of Plastics",
             industry = "Percent of Employment in Industry Sector",
             service = "Percent of Employment in Service Sector",
             ag = "Percent of Employment in Agriculture Sector",
             largest_sector = "Largest Employment Sector")
```
:::


#### Row {height="50%"}

::: panel-tabset
## GDP per Capita Groups

```{r}
#| label: fig-Vizualization-3
#| fig-cap: "Ridge Plot Showing the Amount of Plastic Waste for Each Type of Plastic"

final_merged |> 
  filter(year == 2019) |> 
  group_by(country) |> 
  summarize(unknown = sum(empty),
            hdpe = sum(hdpe),
            ldpe = sum(ldpe),
            other = sum(o),
            pet = sum(pet),
            pp = sum(pp),
            ps = sum(ps),
            pvc = sum(pvc),
            total = sum(total),
            gdp_per_capita = mean(gdp_per_capita),
            hdi = mean(HDI),
            .groups = "drop") |> 
  mutate(income_group = case_when(
    gdp_per_capita < 1500 ~ "Low Income",
    gdp_per_capita > 12000 ~ "High Income",
    TRUE ~ "Middle Income")) |> 
  pivot_longer(
    cols = unknown:pvc,
    names_to = "plastic_type",
    values_to = "waste") |> 
  filter(waste <= 3500) |> 
  
  ggplot(mapping = aes(x = waste,
                       y = plastic_type,
                       fill = income_group)) +
  geom_density_ridges(alpha = 0.7) +
  scale_fill_manual(values = c(
  "High Income"   = "#58b4f5",
  "Middle Income" = "#a8d8ea",
  "Low Income"    = "#56e8c8")) +
  labs(
    title = "Plastic Waste Type Distribution of Countries by GDP per Capita",
    subtitle = "Divided into <span style='color: #58b4f5;'>High Income</span>, 
      <span style='color: #a8d8ea;'>Middle Income</span>,
    and <span style='color: #56e8c8;'>Low Income</span>",
    x = "Waste Amount",
    y = "",
    alt = "Density ridge plot showing the density of plastic waste audited
    for multiple types of plastic. Colored by GDP per capita category,
    High, Middle, and Low") +
  theme_dark() +
  theme(plot.subtitle =  element_markdown(),
        legend.position = "none") 

```

## GDP vs Plastic Pollution

```{r}
#| label: fig-Vizualization-2
#| fig-cap: "Bubble Plot of Plastic Pollution From Audits Against GDP for 2019 and 2020"
#| warning: false

final_merged |>
  mutate(region = countrycode::countrycode(country,
            origin = "country.name",
            destination = "continent"),
         total_normalized = total / num_events, 
         gdp_trillions = gdp / 1e6) |>
  filter_out(country == "EMPTY",
           total == 0,
           is.na(total),
           is.na(gdp_trillions),
           gdp_trillions == 0, 
           is.na(region)) |>
  group_by(country, region, year) |>
  summarize(total_plastic = sum(total_normalized, na.rm = T), 
            gdp_trillions = mean(gdp_trillions, na.rm = T), 
            Population = mean(Population, na.rm = T)) |>
  ggplot(aes(x = gdp_trillions, y = total_plastic, size = Population, 
             color = region)) +
    geom_point(alpha = 0.7) +
    scale_size(range = c(.1, 15), name ="Population") +
  labs(color = "Region",
       x = "GDP (Trillions)", 
       title = "Plastic Pollution Found in Audits vs. GDP for 2019 and 2020",
       subtitle = "Total Plastic per Cleanup Event",
       y = "",
       alt = "Bubble chart showing plastic items found per cleanup event vs GDP in trillions,
       colored by region and sized by population") +
  coord_cartesian(ylim = c(0, 26000),
                  xlim = c(0,25)) + 
  scale_color_discrete(na.translate = FALSE) +
  scale_x_sqrt() +
  facet_wrap(~ year) +
  theme_light() + 
  scale_color_manual(values = c(
  "Africa"   = "#E69F00", 
  "Americas" = "#56B4E9",
  "Asia"     = "#009E73", 
  "Europe"   = "#F0E442", 
  "Oceania"  = "#0072B2"
))

```
:::




```{r}
the_7 <- final_merged |> 
  filter(!parent_company %in% c("Grand Total", 
                                "Unbranded", 
                                "null", 
                                "NULL", 
                                "Assorted")) |> 
  group_by(country, parent_company) |> 
  summarize(total = sum(total), .groups = "drop") |> 
  group_by(country) |> 
  slice_max(order_by = total, n = 7, with_ties = FALSE) |> 
  ungroup()
```




```{r}
#| execute: false
#| include: false

the_7 |> 
  filter(country == "Argentina") |> 

ggplot(mapping = aes(x = reorder(parent_company, total), y = total)) +
  geom_col() +
  labs(title = "Top 7 Parent Companies in Argentina",
       x = "Parent Company",
       y = "Total") +
  theme_minimal()
```













```{r}
library(tidyverse)
library(plotly)
library(MASS)

countries <- unique(the_7$country)

fig <- plot_ly()

for (cty in countries) {

  df <- the_7 |>
    filter(country == cty) |>
    arrange(total) |>
    mutate(parent_company = factor(parent_company, levels = parent_company))

  fig <- fig |>
    add_bars(data = df,
             x = ~parent_company,
             y = ~total,
             name = cty,
             visible = FALSE)
}

buttons <- lapply(seq_along(countries), function(i) {

  df <- the_7 |>
    filter(country == countries[i]) |>
    arrange(total)

  vis <- rep(FALSE, length(countries))
  vis[i] <- TRUE

  list(method = "update",
       args = list(list(visible = vis),
                   list(title = paste("Top 7 Parent Companies in",
                                      countries[i]),
                        xaxis = list(tickmode = "array",
                                     tickvals = 0:nrow(df),
                                     ticktext = df$parent_company))),
       label = countries[i])})

fig <- fig |>
  layout(updatemenus = list(
    list(type = "dropdown",buttons = buttons)),
    title = paste("Top 7 Plastic Waste Producing Companies in", countries[1]),
    xaxis = list(title = "Parent Company"),
    yaxis = list(title = "Total Plastic Audited"),
    showlegend = FALSE)

fig
```







### Column - Model {width="50%"}


```{r}
#| label: Data Model Output1
#| eval: false
#| warning: false
#| include: false

model_df <- final_merged |>
  group_by(country, year) |>
  summarise(
    total_plastic = sum(total, na.rm = TRUE),
    Population_millions = first(na.omit(Population)) / 1000000,
    pop_density = first(na.omit(pop_density)),
    gdp_billions = first(na.omit(gdp)) / 1000000000,
    region = first(na.omit(region)),
    .groups = "drop"
  ) |>
  mutate(
    region = as.factor(region),
    log_total_plastic = log(total_plastic)
  ) 

fit_6 <- lm(
  log_total_plastic ~ Population_millions + pop_density + gdp_billions + region,
  model_df
)

tbl <- tidy(fit_6, conf.int = TRUE) |>
  mutate(
    Predictor = dplyr::recode(term,
      `(Intercept)` = "Intercept",
      `Population_millions` = "Population (millions)",
      `pop_density` = "Population Density",
      `gdp_billions` = "GDP (billions)"
    ),
    Predictor = sub("^region", "Region: ", Predictor)
  ) |>
  select(
    Predictor,
    Estimate = estimate,
    `Std. Error` = std.error,
    `t value` = statistic,
    `p value` = p.value,
    `95% CI Lower` = conf.low,
    `95% CI Upper` = conf.high
  )

tbl |>
  gt() |>
  tab_header(
    title = md("**Linear Regression Predicting Logged Plastic Collected by Country, Year**"),
    subtitle = "Response: log(total_plastic)"
  ) |>
  cols_label(
    Predictor = md("**Predictor**"),
    Estimate = md("**Estimate**"),
    `Std. Error` = md("**Std. Error**"),
    `t value` = md("**t value**"),
    `p value` = md("**p value**"),
    `95% CI Lower` = md("**95% CI Lower**"),
    `95% CI Upper` = md("**95% CI Upper**")
  ) |>
  fmt_number(
    columns = c(Estimate, `Std. Error`, `t value`, `95% CI Lower`, `95% CI Upper`),
    decimals = 4
  ) |>
  fmt_number(columns = `p value`, decimals = 3) |>
  tab_source_note(
    source_note = md(
       "The response variable was transformed as log(total_plastic) to meet linear regression assumptions"
    )
  )
```


# Secondary Question