---
title: "Environmental Plastic and Its Drivers"
format: 
  dashboard:
    orientation: columns
logo: plastics.jpeg
nav-buttons:
      - icon: github
        href: https://github.com/jonflores27/stat541-project-2
        aria-label: GitHub
        
---

```{r}
#| label: Libraries
#| message: false

library(tidyverse)
library(gt)
library(httr)
library(stringr)
library(tidyjson)
library(jsonlite)
library(ggridges)
library(RColorBrewer)
library(viridis)
library(ggtext)
library(broom)
library(scales)
library(car)

plastics <- readr::read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/main/data/2021/2021-01-26/plastics.csv')

plastics_clean <- plastics |> 
  mutate(across(empty:pvc, ~replace_na(., 0)),
         total = rowSums(across(c(empty:pvc))),
         .before = num_events)

final_merged <- readRDS("full_table")
```

## Col
### Row
:::panel-tabset

## Plastic Type Proportions

```{r}
#| label: Data-Summary-1

plastic_cols <- c("hdpe", "ldpe", "o", "pet", "pp", "ps", "pvc")

map(plastic_cols, function(col) {
  tibble(
    plastic_type = col,
    total = sum(plastics[[col]], na.rm = TRUE),
    mean = mean(plastics[[col]], na.rm = TRUE),
    median = median(plastics[[col]], na.rm = TRUE),
    sd = sd(plastics[[col]], na.rm = TRUE),
    iqr = IQR(plastics[[col]], na.rm = TRUE),
    min = min(plastics[[col]], na.rm = TRUE),
    max = max(plastics[[col]], na.rm = TRUE),
    audits_with_plastic_type = sum(plastics[[col]] > 0, na.rm = TRUE),
    percent_of_audits = mean(plastics[[col]] > 0, na.rm = TRUE)
  )
}) |>
  list_rbind() |>
  gt() |>
  fmt_percent(columns = percent_of_audits, decimals = 2) |>
  fmt_number(columns = mean:sd, decimals = 2) |>
  tab_style(
    style = cell_text(weight = "bold"),
    locations = cells_column_labels()
  ) |>
  tab_header("Summary statistics for each plastic type across all audits") |>
  cols_label(
    plastic_type = "Plastic Type",
    total = "Total",
    mean = "Mean",
    median = "Median",
    sd = "SD",
    iqr = "IQR",
    min = "Min",
    max = "Max",
    audits_with_plastic_type = "Audits With Plastic Type",
    percent_of_audits = "Percent of Audits"
  ) |>
  data_color(
    method = "numeric",
    palette = "PuBu", 
    reverse = FALSE
  )
```


## Audit Statistics
```{r}
#| label: Data-Summary-2

country_profile <- function(data, group_var1, group_var2, min_audits = 10) {
  data |>
    filter(!is.na(country), country != "EMPTY") |>
    group_by({{ group_var1 }}, {{ group_var2 }}) |>
    summarise(
      audits = n(),
      total_pieces = sum(grand_total, na.rm = TRUE),
      median_pieces = median(grand_total, na.rm = TRUE),
      mean_pieces = mean(grand_total, na.rm = TRUE),
      max_pieces = max(grand_total, na.rm = TRUE),
      .groups = "drop"
    ) |>
    filter(audits >= min_audits) |>
    arrange({{ group_var1 }}, {{ group_var2 }}) |>
    mutate(country = str_to_title(country)) |>
    filter(country %in% c("India", "China", "United States Of America",
                        "Indonesia", "Nigeria")) |>
    gt() |> 
    tab_header(
      title = md(
        paste0(
          "**Table.** Plastic cleanup profile by country and year for countries 
          with at least ",
           min_audits,
          " audits."
        )
      )
  ) |>
  fmt_number(
    columns = c(audits, total_pieces, median_pieces, mean_pieces, max_pieces),
    decimals = 1,
    use_seps = TRUE
  ) |>
  cols_label(
    country = md("**Country**"),
    year = md("**Year**"),
    audits = md("**Audits**"),
    total_pieces = md("**Total Pieces**"),
    median_pieces = md("**Median Pieces**"),
    mean_pieces = md("**Mean Pieces**"),
    max_pieces = md("**Max Pieces**")
  )
}


country_profile(plastics_clean, country, year, min_audits = 1)
```
:::

### Row
:::panel-tabset

## Waste per Audit
```{r}
#| label: Vizualization-1
#| fig-alt: "Lollipop plot comparing average waste per audit for India, China, the United States, Indonesia, and Nigeria in 2019 and 2020. Each country appears on the y-axis, average waste per audit appears on the x-axis, and separate panels compare the two years."
rate_ify <- function(df, numerator, denominator, name = rate){
  stopifnot(is.data.frame(df))
  df |>
    mutate({{name}} := {{numerator}}/{{denominator}})
}

plastics_clean |> 
  group_by(country, year) |> 
  summarise(total_trash = sum(total),
            num_events = mean(num_events),
            .groups = "drop") |>
  rate_ify(numerator = total_trash,
           denominator = num_events,
           name = avg_event_impact) |> 
  mutate(country = str_to_title(country)) |>
  filter(country %in% c("India",
                        "China",
                        "United States Of America",
                        "Indonesia", 
                        "Nigeria")) |>
  mutate(country = fct_reorder(country, avg_event_impact)) |>
  ggplot(aes(x = avg_event_impact, y = country)) +
  geom_segment(aes(x = 0, xend = avg_event_impact,
                   y = country, yend = country),
               color = "gray60") +
  geom_point(size = 4, color = "steelblue") +
  facet_wrap(~ year) +
  labs(
    title = "Average Waste Accounted for per Audit",
    subtitle = "Top 5 Most Populated Countries",
    x = "Average Waste per Audit",
    y = "",
    alt = "Lollipop plot of the average waste per audit for 2019 and 2020 for the 5 most populated countries, India, China, USA, Indonesia, and Nigeria."
  ) +
  geom_text(aes(label = round(avg_event_impact, 1)), 
          hjust = -0.3, 
          size = 3) +
  theme_minimal()
```

## Plastic Types by Countries
```{r}
#| label: Visualization-2
#| fig-alt: "Bar charts showing the total count of each plastic type collected in audits for India, China, the United States, Indonesia, and Nigeria. Each panel represents one country, and bars compare plastic types such as HDPE, LDPE, PET, PP, PS, PVC, and other plastics."
plastics_clean |>
  filter(country %in% c("India", "China", "United States of America",
                        "Indonesia", "Nigeria")) |>
  mutate(country = str_to_title(country)) |>
  pivot_longer(cols = c(hdpe, ldpe, o, pet, pp, ps, pvc),
               names_to = "plastic_type",
               values_to = "count") |>
  group_by(country, plastic_type) |>
  summarise(total = sum(count, na.rm = TRUE), .groups = "drop") |>
  mutate(plastic_type = fct_reorder(plastic_type, total)) |>
  ggplot(aes(x = total, y = plastic_type, fill = plastic_type)) +
  geom_col(show.legend = FALSE) +
  facet_wrap(~ country, scales = "free_x") +
  scale_fill_brewer(palette = "Blues") +
  labs(
    title = "Most Abundant Plastic Types by 5 Most Populated Countries",
    x = "Total Count",
    y = "",
    alt = "Bar graph of the amount of each plastic type found in audits from the 5 most populated countries, India, China, USA, Indonesia, and Nigeria."
  ) +
  theme_dark()
```

:::


## Col
### Row
:::panel-tabset
## Environmental Metrics
```{r}
#| label: Table Reserch Question 2

final_merged |>
  group_by(country, year) |>
  summarise(
    total_plastic = sum(total),
    avg_plastic = mean(total),
    footprint_per_capita = first(EFConsPerCap[!is.na(EFConsPerCap)]),
    biocapacity_per_capita = first(BiocapPerCap[!is.na(BiocapPerCap)]),
    hdi = first(HDI[!is.na(HDI)]),
    population = first(Population[!is.na(Population)]),
    .groups = "drop"
  ) |>
  filter_out(country == "EMPTY") |>
  filter(country %in% c("India",
                        "China",
                        "United States of America",
                        "Indonesia", 
                        "Nigeria")) |> 
  mutate(country = str_to_title(country)) |> 
  gt() |>
  tab_header(
    title = "Country-Year Summary of Plastics and Environmental Data"
  ) |>
  cols_label(
    country = "Country",
    year = "Year",
    total_plastic = "Total Plastic Items Collected",
    avg_plastic = "Average Plastic Items Collected",
    footprint_per_capita = "Ecological Footprint per Capita",
    biocapacity_per_capita = "Biocapacity per Capita",
    hdi = "HDI",
    population = "Population"
  ) |>
  fmt_number(
    columns = c(total_plastic, avg_plastic, population),
    decimals = 0
  ) |>
  fmt_number(
    columns = c( footprint_per_capita, biocapacity_per_capita, hdi),
    decimals = 2
  )
```

## Employment Sectors
```{r}
#| label: Research question 2 table 2


rate_ify2 <- function(df, numerators, denominator) {
  stopifnot(is.data.frame(df))
  
  df |>
    mutate(across(.cols = {{ numerators }},
                  .fns = ~ .x / {{ denominator }},
                  .names = "{.col}"))
}

final_merged |> 
  group_by(country) |> 
  summarise(hdpe = sum(hdpe),
            ldpe = sum(ldpe),
            o = sum(o),
            pet = sum(pet),
            pp = sum(pp),
            ps = sum(ps),
            pvc = sum(pvc),
            total_waste = sum(total),
            industry = mean(employment_industry),
            service = mean(employment_services),
            ag = mean(employment_agriculture),
            .groups = "drop") |>
  filter(country %in% c("India",
                        "China",
                        "United States of America",
                        "Indonesia", 
                        "Nigeria")) |>
  rate_ify2(numerator = hdpe:pvc,
           denominator = total_waste) |> 

  gt() |> 
  tab_header(
    title = "Proportions of Plastic Types Compared to Employment Sectorization",
    subtitle = "Data from 2019 and 2020 Featuring the 5 Most Populated Countries") |> 
  fmt_percent(
    columns = hdpe:pvc,
    decimals = 2) |> 
  tab_style(style = cell_text(weight = "bold"),
            locations = cells_column_labels()) |>
  cols_label(country = "Country",
             o = "Other",
             total_waste = "Total Count of Plastics",
             industry = "Percent of Employment in Industry Sector",
             service = "Percent of Employment in Service Sector",
             ag = "Percent of Employment in Agriculture Sector") |> 
  data_color(columns = hdpe:pvc,
             direction = "row",
             palette = "PuBu"
             ) |> 
  data_color(columns = industry: ag,
             direction = "row",
             palette = "PuBu"
             )
```
:::

### Row
:::panel-tabset
## Waste Type Distribution by GDP

Teams sorted in descending order of total medals.

```{r}
#| label: Vizualization-3
#| fig-alt: "Density ridge plot showing distributions of audited plastic waste amounts by plastic type for 2019. Ridges are grouped by plastic type, and color indicates GDP per capita income group: low income, middle income, or high income."
final_merged |> 
  filter(year == 2019) |> 
  group_by(country) |> 
  summarize(unknown = sum(empty),
            hdpe = sum(hdpe),
            ldpe = sum(ldpe),
            other = sum(o),
            pet = sum(pet),
            pp = sum(pp),
            ps = sum(ps),
            pvc = sum(pvc),
            total = sum(total),
            gdp_per_capita = mean(gdp_per_capita),
            hdi = mean(HDI),
            .groups = "drop") |> 
  mutate(income_group = case_when(
    gdp_per_capita < 1500 ~ "Low Income",
    gdp_per_capita > 12000 ~ "High Income",
    TRUE ~ "Middle Income")) |> 
  pivot_longer(
    cols = unknown:pvc,
    names_to = "plastic_type",
    values_to = "waste") |> 
  filter(waste <= 3500) |> 
  
  ggplot(mapping = aes(x = waste,
                       y = plastic_type,
                       fill = income_group)) +
  geom_density_ridges(alpha = 0.7) +
  scale_fill_manual(values = c(
  "High Income"   = "#58b4f5",
  "Middle Income" = "#a8d8ea",
  "Low Income"    = "#56e8c8")) +
  labs(
    title = "Plastic Waste Type Distribution of Countries by GDP per Capita",
    subtitle = "Divided into <span style='color: #58b4f5;'>High Income</span>, 
      <span style='color: #a8d8ea;'>Middle Income</span>,
    and <span style='color: #56e8c8;'>Low Income</span>",
    x = "Waste Amount",
    y = "",
    alt = "Density ridge plot showing the density of plastic waste audited
    for multiple types of plastic. Colored by GDP per capita category,
    High, Middle, and Low") +
  theme_dark() +
  theme(plot.subtitle =  element_markdown(),
        legend.position = "none") 

```

## Plastic Type Count by GDP

```{r}
#| label: Vizualization-2
#| fig-alt: "Bubble plot comparing plastic collected per cleanup event with GDP for countries in 2019 and 2020. Each point represents a country-year, the x-axis shows GDP in trillions, the y-axis shows total plastic per cleanup event, point size represents population, color represents region, and panels separate the two years."
#| warning: false

final_merged |>
  mutate(region = countrycode::countrycode(country,
            origin = "country.name",
            destination = "continent"),
         total_normalized = total / num_events, 
         gdp_trillions = gdp / 1e6) |>
  filter_out(country == "EMPTY",
           total == 0,
           is.na(total),
           is.na(gdp_trillions),
           gdp_trillions == 0, 
           is.na(region)) |>
  group_by(country, region, year) |>
  summarize(total_plastic = sum(total_normalized, na.rm = T), 
            gdp_trillions = mean(gdp_trillions, na.rm = T), 
            Population = mean(Population, na.rm = T)) |>
  ggplot(aes(x = gdp_trillions, y = total_plastic, size = Population, 
             color = region)) +
    geom_point(alpha = 0.7) +
    scale_size(range = c(.1, 15), name ="Population") +
  labs(color = "Region",
       x = "GDP (Trillions)", 
       title = "Plastic Pollution Found in Audits vs. GDP for 2019 and 2020",
       subtitle = "Total Plastic per Cleanup Event",
       y = "",
       alt = "Bubble chart showing plastic items found per cleanup event vs GDP in trillions,
       colored by region and sized by population") +
  coord_cartesian(ylim = c(0, 26000),
                  xlim = c(0,25)) + 
  scale_color_discrete(na.translate = FALSE) +
  scale_x_sqrt() +
  facet_wrap(~ year) +
  theme_light() + 
  scale_color_manual(values = c(
  "Africa"   = "#E69F00", 
  "Americas" = "#56B4E9",
  "Asia"     = "#009E73", 
  "Europe"   = "#F0E442", 
  "Oceania"  = "#0072B2"
))
```

:::

## Col
### Row {height="30%"}

:::{.card, title="Introduction}

Reserach Questions:

Other stuff, the why
:::

### Row
```{r}
#| label: Data Model
#| include: false

model_df <- final_merged |>
  group_by(country, year) |>
  summarise(
    total_plastic = sum(total, na.rm = TRUE),
    Population = first(na.omit(Population)),
    pop_density = first(na.omit(pop_density)),
    employment_industry = first(na.omit(employment_industry)),
    HDI = first(na.omit(HDI)),
    gdp = first(na.omit(gdp)),
    imports = first(na.omit(imports)),
    infant_mortality = first(na.omit(infant_mortality)),
    region = first(na.omit(region)),
    .groups = "drop"
  ) |>
  mutate(
    region = as.factor(region),
    log_total_plastic = log(total_plastic)
  ) 

fit_1 <- lm(
  log_total_plastic ~ Population + pop_density + region,
  model_df
)

fit_2 <- lm(
  log_total_plastic ~ Population + pop_density + employment_industry + region,
  model_df
)

fit_3 <- lm(
  log_total_plastic ~ Population + pop_density + HDI + region,
  model_df
)

fit_4 <- lm(
  log_total_plastic ~ Population + pop_density + imports + region,
  model_df
)

fit_5 <- lm(
  log_total_plastic ~ Population + pop_density + infant_mortality + region,
  model_df
)


fit_6 <- lm(
  log_total_plastic ~ Population + pop_density + region + gdp,
  model_df
)

fit_7 <- lm(
  log_total_plastic ~ Population + pop_density + region + gdp + imports,
  model_df
)

adj_rsq <- list(
  summary(fit_1)$adj.r.squared,
  summary(fit_2)$adj.r.squared,
  summary(fit_3)$adj.r.squared,
  summary(fit_4)$adj.r.squared,
  summary(fit_5)$adj.r.squared,
  summary(fit_6)$adj.r.squared,
  summary(fit_7)$adj.r.squared
)

adj_rsq


vif_results <- list(
  vif(fit_1),
  vif(fit_2),
  vif(fit_3),
  vif(fit_4),
  vif(fit_5),
  vif(fit_6),
  vif(fit_7)
)


```

```{r}
#| label: Data Model Output
#| warning: false

model_df <- final_merged |>
  group_by(country, year) |>
  summarise(
    total_plastic = sum(total, na.rm = TRUE),
    Population_millions = first(na.omit(Population)) / 1000000,
    pop_density = first(na.omit(pop_density)),
    gdp_billions = first(na.omit(gdp)) / 1000000000,
    region = first(na.omit(region)),
    .groups = "drop"
  ) |>
  mutate(
    region = as.factor(region),
    log_total_plastic = log(total_plastic)
  ) 

fit_6 <- lm(
  log_total_plastic ~ Population_millions + pop_density + gdp_billions + region,
  model_df
)

tbl <- tidy(fit_6, conf.int = TRUE) |>
  mutate(
    Predictor = dplyr::recode(term,
      `(Intercept)` = "Intercept",
      `Population_millions` = "Population (millions)",
      `pop_density` = "Population Density",
      `gdp_billions` = "GDP (billions)"
    ),
    Predictor = sub("^region", "Region: ", Predictor)
  ) |>
  select(
    Predictor,
    Estimate = estimate,
    `Std. Error` = std.error,
    `t value` = statistic,
    `p value` = p.value,
    `95% CI Lower` = conf.low,
    `95% CI Upper` = conf.high
  )

tbl |>
  gt() |>
  tab_header(
    title = md("**Linear Regression Predicting Logged Plastic Collected by Country, Year**"),
    subtitle = "Response: log(total_plastic)"
  ) |>
  cols_label(
    Predictor = md("**Predictor**"),
    Estimate = md("**Estimate**"),
    `Std. Error` = md("**Std. Error**"),
    `t value` = md("**t value**"),
    `p value` = md("**p value**"),
    `95% CI Lower` = md("**95% CI Lower**"),
    `95% CI Upper` = md("**95% CI Upper**")
  ) |>
  fmt_number(
    columns = c(Estimate, `Std. Error`, `t value`, `95% CI Lower`, `95% CI Upper`),
    decimals = 4
  ) |>
  fmt_number(columns = `p value`, decimals = 3) |>
  tab_source_note(
    source_note = md(
       "The response variable was transformed as log(total_plastic) to meet linear regression assumptions"
    )
  )
```

### Row {height="20%"}

:::{.card, title="conclusions"}

Concluding statement and overall reasons

:::